---
title: "habshd_individual_prs"
subtitle: "Which individual PRSs associate with cognitive impairment?"
format: html
---
## Setup

main input 
master_rds <- "/wynton/group/andrews/users/rakshyasharma/PRS/shared_data/habshd_master_df.rds"

required input columns 
IID
cognitive_impairment
ID_Gender
Age
Race
APOE4_Positivity
PC1, PC2, PC3, PC4

Medication variables:
taking_cholesterol_meds
taking_diabetes_meds
taking_hypertension_meds

Clinical / phenotype variables:
BW_CholTotal
BW_LDLchol
BW_HDLChol
BW_Triglycerides
BW_HbA1c
OM_BMI
OM_BP_DIA
OM_BP_SYS
Pulse_Pressure
CDX_Depression
CDX_Hypertension
CDX_Diabetes
AUDIT_Total
SocialSupport_Total
IMH_TBI
Smoke
Vision_Loss
Hearing_Loss
GDS_Category

PRS variables:
bmi_score
education_score
alcohol_drinking_weekly_score
smoking_initiation_score
diabetes_score
diastolic_blood_pressure_score
systolic_blood_pressure_score
pulse_pressure_score
total_cholesterol_score
social_isolation_score
Hearing_Loss_score
Vision_Loss_score
Triglycerides_score
HDL_Cholesterol_score
LDL_Cholesterol_score
Physical_Activity_score
Major_Depressive_Disorder_score

cognitive_impairment ~ PRS + Age + APOE4_Positivity + PC1 + PC2 + PC3 + PC4 + Race

(ID_gender) for total model

output

results/
├── rds/
│   ├── habshd_individual_prs_analytic_dataset.rds
│   ├── habshd_individual_prs_results.rds
│   └── habshd_individual_prs_sample_sizes.rds
│
├── tables/
│   ├── habshd_individual_prs_baseline_table1.docx
│   ├── habshd_individual_prs_baseline_table1.rtf
│   ├── habshd_individual_prs_baseline_table1.html
│   ├── habshd_individual_prs_results.csv
│   ├── habshd_individual_prs_results.html
│   ├── habshd_individual_prs_sample_sizes.csv
│   ├── habshd_individual_prs_sample_sizes.html
│   └── habshd_individual_prs_results.xlsx
│
└── figures/
    ├── habshd_individual_prs_total.png
    └── habshd_individual_prs_by_sex.png

```{r setup}
.libPaths("/wynton/home/andrews/rakshyasharma/R/rocky8-x86_64-pc-linux-gnu-library/4.4-CBI-gcc13/")

library(readr)
library(dplyr)
library(tidyr)
library(purrr)
library(broom)
library(ggplot2)
library(gt)
library(gtsummary)
library(data.table)
library(readxl)
library(stringr)
library(flextable)
library(openxlsx)

options(dplyr.summarise.inform = FALSE)
theme_set(theme_minimal(base_size = 12, base_family = "Arial"))
```
## Inputs


```{r inputs}
master_rds <- "/wynton/group/andrews/users/rakshyasharma/PRS/shared_data/habshd_master_df.rds"

out_dir <- "results"
out_fig_dir <- file.path(out_dir, "figures")
out_tbl_dir <- file.path(out_dir, "tables")
out_rds_dir <- file.path(out_dir, "rds")

dir.create(out_fig_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(out_tbl_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(out_rds_dir, recursive = TRUE, showWarnings = FALSE)
```

## Helper functions

```{r helpers}
check_required_columns <- function(df, required_cols) {
  missing_cols <- setdiff(required_cols, names(df))
  if (length(missing_cols) > 0) {
    stop(
      paste(
        "The master dataset is missing required columns:",
        paste(missing_cols, collapse = ", ")
      )
    )
  }
}
```
## Load master dataset

```{r analytic-dataset}
final_df <- readRDS(master_rds) %>%
  dplyr::mutate(
    IID = as.character(IID),
    APOE4_Positivity = as.integer(APOE4_Positivity),
    APOE4_Positivity = ifelse(APOE4_Positivity %in% c(0, 1), APOE4_Positivity, NA_integer_),
    Sex = factor(ID_Gender, levels = c(0, 1), labels = c("Male", "Female")),
    RAPA_1_Scores = ifelse(RAPA_1_Total > 6, "High Score", "Low Score"),
    RAPA_2_Scores = ifelse(RAPA_2_Total > 2, "High Score", "Low Score")
  ) %>%
  dplyr::filter(!is.na(APOE4_Positivity))
  
required_cols <- c(
  "IID", "cognitive_impairment", "ID_Gender", "Sex", "Age", "Race",
  "APOE4_Positivity", "PC1", "PC2", "PC3", "PC4",
  "taking_cholesterol_meds", "taking_diabetes_meds", "taking_hypertension_meds",
  "BW_CholTotal", "BW_LDLchol", "BW_HDLChol", "BW_Triglycerides", "BW_HbA1c",
  "OM_BMI", "OM_BP_DIA", "OM_BP_SYS", "Pulse_Pressure",
  "CDX_Depression", "CDX_Hypertension", "CDX_Diabetes",
  "AUDIT_Total", "SocialSupport_Total", "IMH_TBI",
  "Smoke", "Vision_Loss", "Hearing_Loss",
  "RAPA_1_Scores", "RAPA_2_Scores", "GDS_Category",
  "bmi_score", "education_score", "alcohol_drinking_weekly_score",
  "smoking_initiation_score", "diabetes_score",
  "diastolic_blood_pressure_score", "systolic_blood_pressure_score",
  "pulse_pressure_score", "total_cholesterol_score",
  "social_isolation_score", "Hearing_Loss_score", "Vision_Loss_score",
  "Triglycerides_score", "HDL_Cholesterol_score", "LDL_Cholesterol_score",
  "Physical_Activity_score", "Major_Depressive_Disorder_score"
)

check_required_columns(final_df, required_cols)

cat("Analytic dataset rows:", nrow(final_df), "\n")
cat("Analytic dataset columns:", ncol(final_df), "\n")
cat("Cognitive impairment counts:\n")
print(table(final_df$cognitive_impairment, useNA = "ifany"))
cat("Sex by cognitive impairment:\n")
print(table(final_df$Sex, final_df$cognitive_impairment, useNA = "ifany"))
```

## Individual PRS association models

```{r individual-prs-models}
prs_vars <- c(
  "bmi_score",
  "education_score",
  "alcohol_drinking_weekly_score",
  "smoking_initiation_score",
  "diabetes_score",
  "diastolic_blood_pressure_score",
  "systolic_blood_pressure_score",
  "pulse_pressure_score",
  "total_cholesterol_score",
  "social_isolation_score",
  "Hearing_Loss_score",
  "Vision_Loss_score",
  "Triglycerides_score",
  "HDL_Cholesterol_score",
  "LDL_Cholesterol_score",
  "Physical_Activity_score",
  "Major_Depressive_Disorder_score"
)

prs_label_map <- c(
  "bmi_score" = "BMI",
  "education_score" = "Education",
  "alcohol_drinking_weekly_score" = "Alcohol Consumption (weekly)",
  "smoking_initiation_score" = "Smoking Initiation",
  "diabetes_score" = "Type 2 Diabetes",
  "diastolic_blood_pressure_score" = "Diastolic Blood Pressure",
  "systolic_blood_pressure_score" = "Systolic Blood Pressure",
  "pulse_pressure_score" = "Pulse Pressure",
  "total_cholesterol_score" = "Total Cholesterol",
  "social_isolation_score" = "Social Isolation",
  "Hearing_Loss_score" = "Hearing Loss",
  "Vision_Loss_score" = "Vision Loss",
  "Triglycerides_score" = "Triglycerides",
  "HDL_Cholesterol_score" = "HDL Cholesterol",
  "LDL_Cholesterol_score" = "LDL Cholesterol",
  "Physical_Activity_score" = "Physical Activity",
  "Major_Depressive_Disorder_score" = "Major Depressive Disorder"
)

fit_habs_prs <- function(data, prs_var, sex_group = "Total", include_sex = TRUE) {
  covars <- c("Age", "APOE4_Positivity", "PC1", "PC2", "PC3", "PC4", "Race")
  if (include_sex) covars <- c(covars, "ID_Gender")

  form <- as.formula(
    paste("cognitive_impairment ~", paste(c(prs_var, covars), collapse = " + "))
  )

  fit <- glm(form, data = data, family = binomial())

  coef_tab <- summary(fit)$coefficients
  zval <- qnorm(0.975)

  if (!(prs_var %in% rownames(coef_tab))) {
    return(tibble(
      prs_var = prs_var,
      sex_strata = sex_group,
      estimate = NA_real_,
      std.error = NA_real_,
      statistic = NA_real_,
      p.value = NA_real_,
      conf.low = NA_real_,
      conf.high = NA_real_,
      estimate_or = NA_real_,
      conf.low_or = NA_real_,
      conf.high_or = NA_real_
    ))
  }

  beta <- coef_tab[prs_var, "Estimate"]
  se <- coef_tab[prs_var, "Std. Error"]

  tibble(
    prs_var = prs_var,
    sex_strata = sex_group,
    estimate = beta,
    std.error = se,
    statistic = coef_tab[prs_var, "z value"],
    p.value = coef_tab[prs_var, "Pr(>|z|)"],
    conf.low = beta - zval * se,
    conf.high = beta + zval * se,
    estimate_or = exp(beta),
    conf.low_or = exp(beta - zval * se),
    conf.high_or = exp(beta + zval * se)
  )
}

model_df_total <- final_df %>%
  dplyr::filter(!is.na(cognitive_impairment))

model_df_female <- model_df_total %>%
  dplyr::filter(Sex == "Female")

model_df_male <- model_df_total %>%
  dplyr::filter(Sex == "Male")

habs_total_results <- purrr::map_dfr(
  prs_vars,
  ~fit_habs_prs(model_df_total, .x, "Total", TRUE)
)

habs_female_results <- purrr::map_dfr(
  prs_vars,
  ~fit_habs_prs(model_df_female, .x, "Female", FALSE)
)

habs_male_results <- purrr::map_dfr(
  prs_vars,
  ~fit_habs_prs(model_df_male, .x, "Male", FALSE)
)

all_results <- dplyr::bind_rows(
  habs_total_results,
  habs_female_results,
  habs_male_results
) %>%
  dplyr::group_by(sex_strata) %>%
  dplyr::mutate(p_adj_fdr = p.adjust(p.value, method = "fdr")) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(Trait = unname(prs_label_map[prs_var]))

missing_trait_labels <- all_results %>%
  dplyr::filter(is.na(Trait)) %>%
  dplyr::distinct(prs_var)

if (nrow(missing_trait_labels) > 0) {
  print(missing_trait_labels)
  stop("Some prs_var values do not have labels in prs_label_map.")
}

trait_order <- all_results %>%
  dplyr::filter(sex_strata == "Total") %>%
  dplyr::arrange(estimate_or) %>%
  dplyr::pull(Trait)

sample_size_df <- tibble(
  sex_strata = c("Total", "Female", "Male"),
  n_total = c(nrow(model_df_total), nrow(model_df_female), nrow(model_df_male)),
  n_cases = c(
    sum(model_df_total$cognitive_impairment == 1, na.rm = TRUE),
    sum(model_df_female$cognitive_impairment == 1, na.rm = TRUE),
    sum(model_df_male$cognitive_impairment == 1, na.rm = TRUE)
  ),
  n_controls = c(
    sum(model_df_total$cognitive_impairment == 0, na.rm = TRUE),
    sum(model_df_female$cognitive_impairment == 0, na.rm = TRUE),
    sum(model_df_male$cognitive_impairment == 0, na.rm = TRUE)
  )
)

cat("Model rows:", nrow(all_results), "\n")
cat("Sample-size rows:", nrow(sample_size_df), "\n")
```

## Baseline Table 1

```{r baseline-table1}
library(gtsummary)
library(flextable)

baseline_table_data <- model_df_total %>%
  dplyr::mutate(
    taking_cholesterol_meds = factor(taking_cholesterol_meds, levels = c(0, 1), labels = c("No", "Yes")),
    taking_diabetes_meds = factor(taking_diabetes_meds, levels = c(0, 1), labels = c("No", "Yes")),
    taking_hypertension_meds = factor(taking_hypertension_meds, levels = c(0, 1), labels = c("No", "Yes")),
    APOE4 = factor(APOE4_Positivity, levels = c(0, 1), labels = c("Negative", "Positive")),
    Cognitive_Impairment = factor(cognitive_impairment, levels = c(0, 1), labels = c("No", "Yes"))
  ) %>%
  dplyr::select(
    Sex, Age, ID_Education, Cognitive_Impairment, Race, APOE4,
    BW_CholTotal, BW_LDLchol, BW_HDLChol, BW_Triglycerides, BW_HbA1c,
    OM_BMI, OM_BP_DIA, OM_BP_SYS, Pulse_Pressure,
    CDX_Depression, CDX_Hypertension, CDX_Diabetes,
    AUDIT_Total, SocialSupport_Total, IMH_TBI,
    Smoke, Vision_Loss, Hearing_Loss,
    RAPA_1_Scores, RAPA_2_Scores,
    taking_cholesterol_meds, taking_diabetes_meds, taking_hypertension_meds,
    GDS_Category
  )

baseline_table1 <- baseline_table_data %>%
  tbl_summary(
    by = Sex,
    missing = "no",
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} / {N} ({p}%)"
    ),
    label = list(
      ID_Education ~ "Education (years)",
      Cognitive_Impairment ~ "Cognitive impairment",
      BW_CholTotal ~ "Total cholesterol (mg/dL)",
      BW_LDLchol ~ "LDL cholesterol (mg/dL)",
      BW_HDLChol ~ "HDL cholesterol (mg/dL)",
      BW_Triglycerides ~ "Triglycerides (mg/dL)",
      BW_HbA1c ~ "HbA1c",
      OM_BMI ~ "BMI",
      OM_BP_DIA ~ "Diastolic BP (mmHg)",
      OM_BP_SYS ~ "Systolic BP (mmHg)",
      Pulse_Pressure ~ "Pulse pressure (mmHg)",
      CDX_Depression ~ "Clinical depression",
      CDX_Hypertension ~ "Hypertension",
      CDX_Diabetes ~ "Diabetes",
      AUDIT_Total ~ "Alcohol use score",
      SocialSupport_Total ~ "Social support score",
      IMH_TBI ~ "History of TBI",
      Smoke ~ "Smoking status",
      Vision_Loss ~ "Vision loss",
      Hearing_Loss ~ "Hearing loss",
      RAPA_1_Scores ~ "Physical activity score 1",
      RAPA_2_Scores ~ "Physical activity score 2",
      taking_cholesterol_meds ~ "Taking cholesterol meds",
      taking_diabetes_meds ~ "Taking diabetes meds",
      taking_hypertension_meds ~ "Taking hypertension meds",
      GDS_Category ~ "GDS category"
    )
  ) %>%
  add_overall() %>%
  add_p() %>%
  modify_caption("**Table 1. Baseline characteristics of the HABS-HD analytic sample used for individual PRS analyses**")

baseline_table1

baseline_table1_flex <- as_flex_table(baseline_table1)
```



## Results table

```{r results-table}
results_table <- all_results %>%
  dplyr::mutate(
    Trait = factor(Trait, levels = trait_order),
    `Odds Ratio (95% CI)` = sprintf("%.3f (%.3f–%.3f)", estimate_or, conf.low_or, conf.high_or),
    `P Value` = formatC(p.value, format = "e", digits = 2),
    `FDR` = formatC(p_adj_fdr, format = "e", digits = 2)
  ) %>%
  dplyr::arrange(Trait, sex_strata) %>%
  dplyr::mutate(Trait = as.character(Trait)) %>%
  dplyr::select(Trait, sex_strata, `Odds Ratio (95% CI)`, `P Value`, `FDR`)

results_table
```
```{r total-plot}
plot_total <- all_results %>%
  dplyr::filter(sex_strata == "Total") %>%
  dplyr::mutate(
    Trait = factor(Trait, levels = trait_order),
    sig = ifelse(p_adj_fdr < 0.05, "significant", "non-significant")
  )

p_total <- ggplot(
  plot_total,
  aes(
    x = Trait,
    y = estimate_or,
    ymin = conf.low_or,
    ymax = conf.high_or,
    alpha = sig
  )
) +
  geom_pointrange(size = 0.8) +
  geom_errorbar(width = 0.3, linewidth = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "black", linewidth = 0.8) +
  coord_flip() +
  scale_alpha_manual(values = c("significant" = 1, "non-significant" = 0.35)) +
  labs(
    x = "Modifiable Risk Factors",
    y = "Odds Ratio (95% CI)",
    title = "Individual PRS associations with cognitive impairment in HABS-HD"
  ) +
  guides(alpha = "none") +
  theme(
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    plot.title = element_text(size = 14, hjust = 0.5)
  )

p_total
```
```{r sex-plot}
plot_sex <- all_results %>%
  dplyr::filter(sex_strata %in% c("Female", "Male")) %>%
  dplyr::mutate(
    Trait = factor(Trait, levels = trait_order),
    signif = dplyr::case_when(
      p_adj_fdr < 0.001 ~ "***",
      p_adj_fdr < 0.01 ~ "**",
      p_adj_fdr < 0.05 ~ "*",
      TRUE ~ ""
    ),
    star_vjust = if_else(sex_strata == "Female", 2, -1),
    sig = if_else(p_adj_fdr < 0.05, "significant", "non-significant")
  )

p_sex <- ggplot(
  plot_sex,
  aes(
    x = Trait,
    y = estimate_or,
    ymin = conf.low_or,
    ymax = conf.high_or,
    color = sex_strata,
    shape = sex_strata,
    alpha = sig
  )
) +
  geom_pointrange(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(width = 0.3, position = position_dodge(width = 0.6), linewidth = 0.8) +
  geom_text(
    aes(label = signif, vjust = star_vjust),
    position = position_dodge(width = 0.6),
    show.legend = FALSE
  ) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "black", linewidth = 0.8) +
  coord_flip() +
  scale_color_manual(values = c("Female" = "deeppink", "Male" = "blue")) +
  scale_shape_manual(values = c("Female" = 16, "Male" = 17)) +
  scale_alpha_manual(values = c("significant" = 1, "non-significant" = 0.3)) +
  labs(
    x = "Modifiable Risk Factors",
    y = "Odds Ratio (95% CI)",
    title = "Individual PRS associations with cognitive impairment by sex in HABS-HD",
    color = "Sex",
    shape = "Sex"
  ) +
  guides(
    color = guide_legend(override.aes = list(shape = c(16, 17), alpha = 1)),
    shape = "none",
    alpha = "none"
  ) +
  theme(
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 14),
    plot.title = element_text(size = 14, hjust = 0.5),
    legend.position = "bottom",
    legend.direction = "horizontal",
    plot.margin = margin(1, 1, 1, 2, "cm")
  )

p_sex
```
```{r sample-size-table}
sample_size_table <- sample_size_df %>%
  dplyr::arrange(sex_strata)

sample_size_table

sample_size_gt <- sample_size_table %>%
  gt() %>%
  tab_header(
    title = "Table 2. Sample-size summary for HABS-HD individual PRS analyses"
  )

sample_size_gt
results_gt <- results_table %>%
  gt() %>%
  tab_header(
    title = "Table 3. HABS-HD individual PRS association results"
  )

results_gt
```
```{r exports}
flextable::save_as_docx(
  "HABS-HD Baseline Table 1" = baseline_table1_flex,
  path = file.path(out_tbl_dir, "habshd_individual_prs_baseline_table1.docx")
)

gt::gtsave(
  as_gt(baseline_table1),
  filename = file.path(out_tbl_dir, "habshd_individual_prs_baseline_table1.rtf")
)

saveRDS(final_df, file.path(out_rds_dir, "habshd_individual_prs_analytic_dataset.rds"))
saveRDS(all_results, file.path(out_rds_dir, "habshd_individual_prs_results.rds"))
saveRDS(sample_size_df, file.path(out_rds_dir, "habshd_individual_prs_sample_sizes.rds"))

write.csv(results_table, file.path(out_tbl_dir, "habshd_individual_prs_results.csv"), row.names = FALSE)
write.csv(sample_size_table, file.path(out_tbl_dir, "habshd_individual_prs_sample_sizes.csv"), row.names = FALSE)
xlsx_file <- file.path(out_tbl_dir, "habshd_individual_prs_results.xlsx")

wb <- openxlsx::createWorkbook()

openxlsx::addWorksheet(wb, "Baseline_Table1")
openxlsx::writeData(wb, "Baseline_Table1", baseline_table_data)

openxlsx::addWorksheet(wb, "Sample_Sizes")
openxlsx::writeData(wb, "Sample_Sizes", sample_size_table)

openxlsx::addWorksheet(wb, "Main_Results")
openxlsx::writeData(wb, "Main_Results", results_table)

openxlsx::addWorksheet(wb, "Raw_Model_Results")
openxlsx::writeData(wb, "Raw_Model_Results", all_results)

openxlsx::saveWorkbook(wb, xlsx_file, overwrite = TRUE)

cat("Workbook written to:", xlsx_file, "\n")
gt::gtsave(
  sample_size_gt,
  filename = file.path(out_tbl_dir, "habshd_individual_prs_sample_sizes.html")
)
gt::gtsave(
  gt::gt(results_table) %>% 
    tab_header(title = "Table 3. HABS-HD individual PRS association results"),
  filename = file.path(out_tbl_dir, "habshd_individual_prs_results.html")
)

gt::gtsave(
  as_gt(baseline_table1),
  filename = file.path(out_tbl_dir, "habshd_individual_prs_baseline_table1.html")
)

ggsave(
  filename = file.path(out_fig_dir, "habshd_individual_prs_total.png"),
  plot = p_total,
  width = 10,
  height = 6,
  dpi = 300
)

ggsave(
  filename = file.path(out_fig_dir, "habshd_individual_prs_by_sex.png"),
  plot = p_sex,
  width = 12,
  height = 14,
  dpi = 600
)

stopifnot(nrow(all_results) == length(prs_vars) * 3)
stopifnot(nrow(sample_size_df) == 3)

cat("Finished successfully.\n")
```

