---
title: "compositePRS"
format: html
---

This QMD trains composite PRS models in ADGC, compares plain logistic/ridge/elastic-net models, tests whether the ADGC-trained composite PRS predicts cognitive impairment in HABS-HD, tests whether the HABS-HD composite PRS relates to plasma biomarkers, and evaluates which individual PRSs contribute most through leave-one-out analysis.

Loads config.yaml
config controls 
seed
ADGC input paths
HABS-HD input paths
PRS folder names
results directory

R/helpers.R --> package loading, trait labels, PRS variable definitions, config helpers

R/load_data.R --> ADGC/HABS-HD data loading and PRS merging

R/models.R --> model-fitting functions

R/exports.R --> Excel/table export functions

R/plots.R --> plotting functions

1. ADGC PLINK family/sample file:
   /wynton/group/andrews/users/rakshyasharma/PRS/PRS-CS-Auto/resources/ADGC/adgc.fam

2. ADGC covariate file:
   /wynton/group/andrews/data/adgc/ADGCdatasets/covar_covariates.tsv

3. ADGC pgsc_calc PRS result folders:
   /wynton/group/andrews/users/rakshyasharma/pgsc_calc/adgc/results/ADGC
   
Graham2021tc_eur
Graham2021TG_eur
Graham2021HDL_eur
Graham2021LDL_eur
Keaton2024diaBP
Keaton2024PP
Keaton2024sysBP
Okbay2022educ
SaundersDrnkwk
SaundersSmkInt
SuzukiDia2024
YengoBMI2018
Socrates2024
Trpchevska2022
Choquet2021
Klimentidis2018mvpa
Adams2025mdd

reads the already-built shared master file: habshd_master_df <- readRDS(
  "/wynton/group/andrews/users/rakshyasharma/PRS/shared_data/habshd_master_df.rds"
)

train_df --> 70% ADGC training data, cases and controls sampled separately

test_df --> 30% ADGC held-out test data

train_df_F --> female-only ADGC training data

test_df_F --> female-only ADGC test data

train_df_M --> male-only ADGC training data

test_df_M --> male-only ADGC test data


adgc_df_prs.rds --> Full ADGC PRS analysis dataset

habshd_eval.rds --> Filtered HABS-HD external validation dataset

plain_total.rds --> Total-sample plain logistic composite PRS model object

plain_female.rds --> Female-only plain logistic composite PRS model object

plain_male.rds --> Male-only plain logistic composite PRS model object

plain_results.rds --> Plain logistic model summary results

ridge_results.rds --> Ridge model summary results

enet_results.rds --> Elastic-net model summary results

compare_results.rds --> Combined plain/ridge/elastic-net comparison results

hab_results.rds --> HABS-HD external validation results

endo_results.rds --> Composite PRS biomarker/endophenotype results in HABS-HD

loo_adgc_results.rds --> ADGC leave-one-PRS-out results

loo_hab_results.rds --> HABS-HD leave-one-PRS-out results

Composite PRS model results
ADGC train/test model results
Ridge and elastic net comparison results
External HABS-HD validation results
HABS-HD biomarker endophenotype results
Leave-one-PRS-out sensitivity results
Table 1 outputs
Figures
RDS objects
Excel workbook

output tree
results/
├── Composite_PRS_main_results.xlsx
│
├── rds/
│   ├── adgc_df_prs.rds
│   ├── habshd_eval.rds
│   ├── plain_total.rds
│   ├── plain_female.rds
│   ├── plain_male.rds
│   ├── plain_results.rds
│   ├── ridge_results.rds
│   ├── enet_results.rds
│   ├── compare_results.rds
│   ├── hab_results.rds
│   ├── endo_results.rds
│   ├── loo_adgc_results.rds
│   └── loo_hab_results.rds
│
├── tables/
│   ├── ADGC_table1_analytic_sample.docx
│   ├── HABSHD_table1_analytic_sample.docx
│   ├── ADGC_table1_analytic_sample.rtf
│   └── HABSHD_table1_analytic_sample.rtf
│
└── figures/
    ├── adgc_model_comparison.png
    ├── habs_main_results.png
    ├── endophenotypes.png
    ├── loo_adgc.png
    └── loo_hab.png
    
    
    
flow chart 

config.yml + R helper scripts
        |
        v
Load ADGC sample/covariate/PRS data
        |
        v
adgc_df_prs
        |
        v
70/30 stratified train-test split
        |
        v
Train composite PRS models in ADGC
        |
        |-- Plain logistic
        |-- Ridge
        |-- Elastic net
        |
        v
ADGC held-out model comparison
        |
        v
compare_results
        |
        v
Read HABS-HD master dataset
        |
        v
habshd_eval
        |
        v
Apply ADGC-trained composite PRS to HABS-HD
        |
        v
hab_results
        |
        v
Run composite PRS biomarker models in HABS-HD
        |
        v
endo_results
        |
        v
Run leave-one-PRS-out analyses
        |
        |-- loo_adgc_results
        |-- loo_hab_results
        |
        v
Export RDS files, Excel workbook, Table 1 files, and figures


```{r setup}

target_lib <- "/wynton/home/andrews/rakshyasharma/R/rocky8-x86_64-pc-linux-gnu-library/4.4-CBI-gcc13"

.libPaths(c(target_lib, .libPaths()))

cat("R version:", R.version.string, "\n")
cat("Library paths used by compositePRS.qmd:\n")
print(.libPaths())
cat("openxlsx available:", requireNamespace("openxlsx", quietly = TRUE), "\n")

source("R/helpers.R")
source("R/load_data.R")
source("R/models.R")
source("R/exports.R")
source("R/plots.R")

load_required_packages()
cfg <- load_project_config("config.yml")

set.seed(cfg$seed)
```

```{r adgc-data}
adgc_df_prs <- load_adgc_data(
  fam_file = cfg$paths$fam_file,
  covar_file = cfg$paths$covar_file,
  base_dir = cfg$paths$adgc_base_dir,
  folder_names = cfg$folder_names,
  trait_labels = trait_labels,
  all_prs_vars = all_prs_vars
)

cat("Rows in ADGC analysis dataset:", nrow(adgc_df_prs), "\n")
cat("Columns in ADGC analysis dataset:", ncol(adgc_df_prs), "\n")
```

```{r train-test-split}
case_ids <- adgc_df_prs %>% dplyr::filter(status_binary == 1) %>% dplyr::pull(ID_2)
control_ids <- adgc_df_prs %>% dplyr::filter(status_binary == 0) %>% dplyr::pull(ID_2)

train_case_ids <- sample(case_ids, size = floor(0.7 * length(case_ids)))
train_control_ids <- sample(control_ids, size = floor(0.7 * length(control_ids)))
train_ids <- c(train_case_ids, train_control_ids)

train_df <- adgc_df_prs %>% dplyr::filter(ID_2 %in% train_ids)
test_df  <- adgc_df_prs %>% dplyr::filter(!ID_2 %in% train_ids)

train_df_F <- train_df %>% dplyr::filter(Sex == 2)
test_df_F  <- test_df %>% dplyr::filter(Sex == 2)

train_df_M <- train_df %>% dplyr::filter(Sex == 1)
test_df_M  <- test_df %>% dplyr::filter(Sex == 1)

cat("Train rows:", nrow(train_df), "\n")
cat("Test rows:", nrow(test_df), "\n")

```

```{r adgc-main-models}
plain_total  <- fit_plain_cprs(train_df,   test_df,   all_prs_vars, "Total",  TRUE)
plain_female <- fit_plain_cprs(train_df_F, test_df_F, all_prs_vars, "Female", FALSE)
plain_male   <- fit_plain_cprs(train_df_M, test_df_M, all_prs_vars, "Male",   FALSE)

plain_results <- dplyr::bind_rows(
  plain_total$result,
  plain_female$result,
  plain_male$result
)

plain_results %>%
  dplyr::select(sex_strata, estimate, conf.low, conf.high, p.value, test_auc)

```

```{r adgc-penalized-models}

ridge_total  <- fit_penalized_cprs(train_df,   test_df,   all_prs_vars, 0,   "Total",  TRUE,  "Ridge")
ridge_female <- fit_penalized_cprs(train_df_F, test_df_F, all_prs_vars, 0,   "Female", FALSE, "Ridge")
ridge_male   <- fit_penalized_cprs(train_df_M, test_df_M, all_prs_vars, 0,   "Male",   FALSE, "Ridge")

enet_total   <- fit_penalized_cprs(train_df,   test_df,   all_prs_vars, 0.5, "Total",  TRUE,  "Elastic net")
enet_female  <- fit_penalized_cprs(train_df_F, test_df_F, all_prs_vars, 0.5, "Female", FALSE, "Elastic net")
enet_male    <- fit_penalized_cprs(train_df_M, test_df_M, all_prs_vars, 0.5, "Male",   FALSE, "Elastic net")

ridge_results <- dplyr::bind_rows(ridge_total$result, ridge_female$result, ridge_male$result)
enet_results  <- dplyr::bind_rows(enet_total$result, enet_female$result, enet_male$result)

compare_results <- dplyr::bind_rows(
  plain_results,
  ridge_results,
  enet_results
) %>%
  dplyr::mutate(
    sex_strata = factor(sex_strata, levels = c("Total", "Male", "Female")),
    model = factor(model, levels = c("Plain logistic", "Ridge", "Elastic net"))
  )

compare_results %>%
  dplyr::select(sex_strata, model, estimate, conf.low, conf.high, p.value, test_auc)

```

```{r habshd-data}

habshd_master_df <- readRDS("/wynton/group/andrews/users/rakshyasharma/PRS/shared_data/habshd_master_df.rds")

habshd_eval <- habshd_master_df %>%
  dplyr::mutate(
    Sex = dplyr::case_when(
      ID_Gender == 0 ~ 1L,  # Male in ADGC coding
      ID_Gender == 1 ~ 2L,  # Female in ADGC coding
      TRUE ~ NA_integer_
    )
  ) %>%
  dplyr::filter(
    !is.na(cognitive_impairment),
    !is.na(Age),
    !is.na(APOE4_Positivity),
    !is.na(Race),
    !is.na(PC1), !is.na(PC2), !is.na(PC3), !is.na(PC4),
    !is.na(ID_Gender),
    !is.na(Sex)
  )
```
```{r habshd-main-results}
hab_total_obj <- score_external_habshd(
  hab_data = habshd_eval,
  model_obj = plain_total,
  outcome_var = "cognitive_impairment",
  sex_subset = NULL,
  sex_group = "Total"
)

hab_female_obj <- score_external_habshd(
  hab_data = habshd_eval,
  model_obj = plain_female,
  outcome_var = "cognitive_impairment",
  sex_subset = 2,
  sex_group = "Female"
)

hab_male_obj <- score_external_habshd(
  hab_data = habshd_eval,
  model_obj = plain_male,
  outcome_var = "cognitive_impairment",
  sex_subset = 1,
  sex_group = "Male"
)

hab_results <- dplyr::bind_rows(
  hab_total_obj$result,
  hab_female_obj$result,
  hab_male_obj$result
)

hab_results %>%
  dplyr::select(group, estimate, conf.low, conf.high, p.value, test_auc)
```

```{r endophenotypes}
hab_total_scored  <- hab_total_obj$data
hab_female_scored <- hab_female_obj$data
hab_male_scored   <- hab_male_obj$data

endo_results <- run_endophenotype_models(
  hab_total_scored = hab_total_scored,
  hab_female_scored = hab_female_scored,
  hab_male_scored = hab_male_scored
)

endo_results %>%
  dplyr::select(group, outcome, estimate, conf.low, conf.high, p.value, p_adj_fdr)

```
```{r loo-analysis}
loo_adgc_total <- run_loo_plain_adgc(train_df,   test_df,   all_prs_vars, include_sex = TRUE,  sex_group = "Total")
loo_adgc_female <- run_loo_plain_adgc(train_df_F, test_df_F, all_prs_vars, include_sex = FALSE, sex_group = "Female")
loo_adgc_male <- run_loo_plain_adgc(train_df_M, test_df_M, all_prs_vars, include_sex = FALSE, sex_group = "Male")

loo_adgc_results <- dplyr::bind_rows(
  loo_adgc_total,
  loo_adgc_female,
  loo_adgc_male
)

loo_hab_total <- run_loo_habshd(
  train_df = train_df,
  test_df = test_df,
  hab_data = habshd_eval,
  prs_vars = all_prs_vars
)

loo_hab_female <- run_loo_habshd_sex(
  train_df = train_df_F,
  test_df = test_df_F,
  hab_data = habshd_eval,
  prs_vars = all_prs_vars,
  sex_code = 2,
  sex_group = "Female"
)

loo_hab_male <- run_loo_habshd_sex(
  train_df = train_df_M,
  test_df = test_df_M,
  hab_data = habshd_eval,
  prs_vars = all_prs_vars,
  sex_code = 1,
  sex_group = "Male"
)

loo_hab_results <- dplyr::bind_rows(
  loo_hab_total,
  loo_hab_female,
  loo_hab_male
)

loo_adgc_results %>% dplyr::slice_head(n = 5)
loo_hab_results %>% dplyr::slice_head(n = 5)

```

```{r plot-ready-objects}
full_adgc_plot <- plain_results %>%
  dplyr::mutate(
    dropped_prs = "Full composite",
    sex_strata = factor(sex_strata, levels = c("Total", "Female", "Male"))
  ) %>%
  dplyr::select(sex_strata, dropped_prs, estimate, conf.low, conf.high, p.value, test_auc)

loo_adgc_plot_df <- dplyr::bind_rows(
  full_adgc_plot,
  loo_adgc_results %>%
    dplyr::mutate(sex_strata = factor(sex_strata, levels = c("Total", "Female", "Male"))) %>%
    dplyr::select(sex_strata, dropped_prs, estimate, conf.low, conf.high, p.value, test_auc)
) %>%
  dplyr::mutate(
    dropped_prs = factor(
      dropped_prs,
      levels = c("Full composite", rev(unique(loo_adgc_results$dropped_prs)))
    )
  )

full_hab_plot <- hab_results %>%
  dplyr::mutate(
    dropped_prs = "Full composite",
    group = factor(group, levels = c("Total", "Female", "Male"))
  ) %>%
  dplyr::select(group, dropped_prs, estimate, conf.low, conf.high, p.value, test_auc)

loo_hab_plot_df <- dplyr::bind_rows(
  full_hab_plot,
  loo_hab_results %>%
    dplyr::mutate(group = factor(group, levels = c("Total", "Female", "Male"))) %>%
    dplyr::select(group, dropped_prs, estimate, conf.low, conf.high, p.value, test_auc)
) %>%
  dplyr::mutate(
    dropped_prs = factor(
      dropped_prs,
      levels = c("Full composite", rev(unique(loo_hab_results$dropped_prs)))
    )
  )

```

```{r save-rds}
dir.create(file.path(cfg$paths$results_dir, "rds"), recursive = TRUE, showWarnings = FALSE)

saveRDS(adgc_df_prs, file.path(cfg$paths$results_dir, "rds", "adgc_df_prs.rds"))
saveRDS(habshd_eval, file.path(cfg$paths$results_dir, "rds", "habshd_eval.rds"))

saveRDS(plain_total, file.path(cfg$paths$results_dir, "rds", "plain_total.rds"))
saveRDS(plain_female, file.path(cfg$paths$results_dir, "rds", "plain_female.rds"))
saveRDS(plain_male, file.path(cfg$paths$results_dir, "rds", "plain_male.rds"))

saveRDS(plain_results, file.path(cfg$paths$results_dir, "rds", "plain_results.rds"))
saveRDS(ridge_results, file.path(cfg$paths$results_dir, "rds", "ridge_results.rds"))
saveRDS(enet_results, file.path(cfg$paths$results_dir, "rds", "enet_results.rds"))
saveRDS(compare_results, file.path(cfg$paths$results_dir, "rds", "compare_results.rds"))
saveRDS(hab_results, file.path(cfg$paths$results_dir, "rds", "hab_results.rds"))
saveRDS(endo_results, file.path(cfg$paths$results_dir, "rds", "endo_results.rds"))
saveRDS(loo_adgc_results, file.path(cfg$paths$results_dir, "rds", "loo_adgc_results.rds"))
saveRDS(loo_hab_results, file.path(cfg$paths$results_dir, "rds", "loo_hab_results.rds"))

```

```{r export-workbook}
xlsx_out <- file.path(cfg$paths$results_dir, "Composite_PRS_main_results.xlsx")

export_main_results_xlsx(
  plain_results = plain_results,
  ridge_results = ridge_results,
  enet_results = enet_results,
  compare_results = compare_results,
  hab_results = hab_results,
  output_file = xlsx_out
)

cat("Workbook written to:", xlsx_out, "\n")

```
## ADGC Table 1

```{r adgc-table1}
library(gtsummary)
library(flextable)
library(gt)

adgc_table1_data <- adgc_df_prs %>%
  dplyr::transmute(
    Sex = factor(Sex, levels = c(1, 2), labels = c("Male", "Female")),
    Status = factor(status, levels = c(1, 2), labels = c("Cognitively Unimpaired", "Alzheimer's Disease")),
    Age = aaoaae,
    APOE4 = factor(apoe4any, levels = c(0, 1), labels = c("Negative", "Positive")),
    Ancestry = MostSimilarPop
  )

adgc_table1_summary <- adgc_table1_data %>%
  tbl_summary(
    by = Sex,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} / {N} ({p}%)"
    ),
    missing = "no"
  ) %>%
  add_overall() %>%
  modify_caption("**Table 1. Baseline characteristics of the ADGC analytic sample**")

adgc_table1_summary

```

## HABS-HD Table 1

```{r habshd-table1}
library(gtsummary)
library(flextable)
library(gt)

habshd_table1_data <- habshd_eval %>%
  dplyr::transmute(
    Sex = factor(ID_Gender, levels = c(0, 1), labels = c("Male", "Female")),
    Cognitive_Impairment = factor(cognitive_impairment, levels = c(0, 1), labels = c("No", "Yes")),
    Age = Age,
    Race = Race,
    APOE4 = factor(APOE4_Positivity, levels = c(0, 1), labels = c("Negative", "Positive"))
  )

habshd_table1_summary <- habshd_table1_data %>%
  tbl_summary(
    by = Sex,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} / {N} ({p}%)"
    ),
    missing = "no"
  ) %>%
  add_overall() %>%
  modify_caption("**Table 2. Baseline characteristics of the HABS-HD analytic sample**")

habshd_table1_summary
```

## Export Table 1 outputs

```{r export-table1}
dir.create(file.path(cfg$paths$results_dir, "tables"), recursive = TRUE, showWarnings = FALSE)

adgc_table1_flex <- as_flex_table(adgc_table1_summary)
habshd_table1_flex <- as_flex_table(habshd_table1_summary)

flextable::save_as_docx(
  "ADGC Table 1" = adgc_table1_flex,
  path = file.path(cfg$paths$results_dir, "tables", "ADGC_table1_analytic_sample.docx")
)

flextable::save_as_docx(
  "HABS-HD Table 1" = habshd_table1_flex,
  path = file.path(cfg$paths$results_dir, "tables", "HABSHD_table1_analytic_sample.docx")
)

gt::gtsave(
  as_gt(adgc_table1_summary),
  filename = file.path(cfg$paths$results_dir, "tables", "ADGC_table1_analytic_sample.rtf")
)

gt::gtsave(
  as_gt(habshd_table1_summary),
  filename = file.path(cfg$paths$results_dir, "tables", "HABSHD_table1_analytic_sample.rtf")
)
```

```{r main-plots}
p_adgc_compare <- plot_adgc_model_comparison(compare_results)
p_hab_main <- plot_hab_main_results(hab_results)
p_endo <- plot_endophenotypes(endo_results)
p_loo_adgc <- plot_loo_adgc(loo_adgc_plot_df)
p_loo_hab <- plot_loo_hab(loo_hab_plot_df)

p_adgc_compare
p_hab_main
p_endo
p_loo_adgc
p_loo_hab
```

```{r save-plots}
dir.create(file.path(cfg$paths$results_dir, "figures"), recursive = TRUE, showWarnings = FALSE)

ggplot2::ggsave(
  filename = file.path(cfg$paths$results_dir, "figures", "adgc_model_comparison.png"),
  plot = p_adgc_compare,
  width = 8,
  height = 6,
  dpi = 300
)

ggplot2::ggsave(
  filename = file.path(cfg$paths$results_dir, "figures", "habs_main_results.png"),
  plot = p_hab_main,
  width = 8,
  height = 6,
  dpi = 300
)

ggplot2::ggsave(
  filename = file.path(cfg$paths$results_dir, "figures", "endophenotypes.png"),
  plot = p_endo,
  width = 9,
  height = 6,
  dpi = 300
)

ggplot2::ggsave(
  filename = file.path(cfg$paths$results_dir, "figures", "loo_adgc.png"),
  plot = p_loo_adgc,
  width = 10,
  height = 8,
  dpi = 300
)

ggplot2::ggsave(
  filename = file.path(cfg$paths$results_dir, "figures", "loo_hab.png"),
  plot = p_loo_hab,
  width = 10,
  height = 8,
  dpi = 300
)

```


## Analytic sample checks

```{r sample-checks}
cat("ADGC analytic N:", nrow(adgc_df_prs), "\n")
cat("HABS-HD analytic N:", nrow(habshd_eval), "\n")

stopifnot(nrow(adgc_df_prs) == nrow(dplyr::distinct(adgc_df_prs, ID_2)))
stopifnot(nrow(habshd_eval) == nrow(dplyr::distinct(habshd_eval, IID)))
stopifnot(length(intersect(train_df$ID_2, test_df$ID_2)) == 0)
stopifnot(nrow(plain_results) == 3)
stopifnot(nrow(hab_results) == 3)
```